Supplementary material for "Crowdsourced geometric morphometrics enable rapid large-scale collection and

### 2015-09-22 manuscript/manuscript.Rmd
Code drop

```{r prelude, echo=FALSE, message=FALSE, warning=FALSE}
library("MASS")
library("readr")
library("dplyr")
library("ggplot2")
library("stringr")
library("magrittr")
library("tidyr")
library("reshape2")
library("scales")
library("geomorph")
library("rjson")
library("ipred")
library("vegan")
library("lubridate")
library("directlabels")
library("BAMMtools")
library("knitr")
library("kfigr")
library("knitcitations")
library("parallel")
library("fish.reliability")
options("citation_format"="pandoc")
cleanbib()
set.seed(111) # reproducible and stable sampling
data(fish_reliability)
data(fish_families)
knitr::opts_chunk$set(echo=FALSE, cache=1, fig.align="center")
```

@@

Although it is now possible to build phylogenetic trees with thousands of tips, and phenotypic data sets have similarly been growing larger and larger, the traits that are typically studied at this scale tend to be simple: geographic occurrences [@Jetz2012], one or two continuous characters [@Harmon2010; @Rabosky2013], a single discrete character [@Goldberg2010; @Edwards2011; @Price2012], or some combination of these [@Zanne2014; @Pyron2014]. A richer understanding of the forces that shape macroevolution requires the collection of more detailed phenotypic trait data at scale.

@@

Collecting landmark-based geometric morphometric data at scale permits detailed analysis of different sources of error, such as among- and within-observer variation [@VonCramon-Taubadel2007]. To assess whether the quality of data gathered by workers recruited through Amazon Mechanical Turk was significantly different than traditionally-collected data, we asked turkers ($n = `r nrow(filter(fish_reliability, role == "turker") %>% distinct(sha1mac))`$) and experts ($n = `r nrow(filter(fish_reliability, role == "morphologist") %>% distinct(sha1mac))`$) to landmark a set of five fish images, five times each. All participants used the same protocol and same software to digitize the same set of fishes. The landmarks were carefully selected based on previously-published literature concerning fish shape [Supplemental Figure S2; @Fink1995; @Cavalcanti1999; @Ruber2001a; @Klingenberg2003; @Chakrabarty2005; @Frederich2008; @Claverie2014; @Thacker2014]. We also ensured that the chosen landmarks included morphological features that were relatively straightforward to digitize (the position of the eye) and features that were likely to be more challenging to digitize (the position of the preopercle bone), in order to test for turker and expert differences over a spectrum of difficulties. We report the inter-observer reliability for turkers and experts by computing the ratio of the among-individual and the sum of the among-individual and measurement error variance components in a repeated measures nested MANOVA [@Palmer1986; @Zelditch2012].

To assess the differences between turker and experts on a per-landmark basis, we first compared the median turker position to the median expert position of each landmark. We assumed that the expert median was the true position of that landmark, and calculated the absolute Euclidian distance. Larger distances would indicate low turker accuracy, while smaller distances would indicate high turker accuracy. We then examined the variance in turker landmarks. For each landmark, we rotated the cloud of points to maximize variance in one dimension, and calculated the log-ratio of median absolute deviations (MAD) between turkers and experts. This rotation is a conservative approach for assessing the difference in variance between these two groups, because it maximizes any apparent differences in landmark position. A positive log-ratio indicated that experts had lower variance than turkers, while a negative log-ratio indicated that turkers had lower variance. For all subsequent analysis, we excluded landmarks where turkers performed especially poorly, where either the accuracy or precision components for a given landmark exceeded 1.5 times the interquartile range of that component.

@@

To demonstrate the utility of obtaining comparative data using this method, we use previously published phylogenies for seven fish families: Acanthuridae [@Sorenson2013], Balistoidae, Tetraodontidae [@Santini2013], Apogonidae, Chaetodontidae, Labridae [@Cowman2011; @Choat2012], and Pomacentridae [@Frederich2013]. We matched `r nrow(distinct(fish_families, tip))` species to left-lateral images from the Encyclopedia of Life (<http://eol.org/>) using their application programming interface [@Parr2014]. Crowdsourced workers placed landmarks describing body shape variation following a standard protocol (Supplementary Material). The Cartesian position of these landmarks were used in a generalized Procrustes analyses [@Gower1975; @Rohlf1990], which centers, scales, and rotates landmark configurations to minimize the least-squares distance between shapes. We then determined the major components of shape variation using a Procrustes-aligned principal components analysis (PCA) [@Mardia1979; @Bookstein1991] with the R package *geomorph* `r citep(citation("geomorph"))`, and used these principal components axes for subsequent analyses.

We used Bayesian Analysis of Macroevolutionary Mixtures [BAMM; @Rabosky2014a] to estimate rates of speciation and body shape evolution for all seven families. For the characters describing body shape, we use the PC axes whose eigenvalues exceeded the corresponding random broken-stick component [@Jackson1993; @Legendre1998]. BAMM estimates the location of rate shifts in either diversification or character evolution using a transdimensional (reversible jump) Markov Chain Monte Carlo method that samples a variety of models of lineage diversification and trait evolution. We assessed convergence and mixing using Tracer [@rambaut2007tracer]. We also repeated each analysis and simulated under the prior (without data) to exclude rate heterogeneity that occurred solely due to stochastic processes. We use a Bayes Factor criterion of $BF > 5$ to enumerate the set of credible shifts [@Shi2015] and visualized them in R using BAMMtools `r citep(citation("BAMMtools"))`.

@@

The inter-observer reliability of turkers and experts as measured by the ratio of among-individual and sum of the among-individual and measurement error ANOVA components was `r percent(reliability$turker)` and `r percent(reliability$expert)`, respectively. Although there is no current standard for acceptable levels of measurement reliability [@VonCramon-Taubadel2007], these percentages are not low enough to suggest pathologies in the measurement protocol.

@@

We projected turker and expert shape configurations into morphospace (Figure `r figr("turk_morph_procrustes_space")`, Supplemental Figure S4) Although the overall space occupied by each family's shape configurations vary, in practice, the aggregated median turker and expert shapes are not qualitatively different. The only exception is the triggerfishes (Balistidae), likely due to turker confusion over the exact location of dorsal fin due to their reduced anterior dorsal fin.

@@

Using a median expert time of `r as.character(morpho_per_image)` per image, we estimate that a single morphologist would take `r as.character(morpho_per_image * nrow(distinct(fish_families, tip)))` to landmark all `r nrow(distinct(fish_families, tip))` images. At 5x replication, this would take `r as.character(morpho_per_image * nrow(fish_families))`. By comparison, turkers took a total of `r as.character(total_turker_time)` to complete all images at 5x replication.

@@

```{r load_bamm_data, message=FALSE, echo=FALSE, warning=FALSE, results="hide"}
base <- "data-raw/bamm/"
base <- "../data-raw/bamm/"
fams <- c("acanthurid", "apogonidae", "balistoidae", "chaetodontidae", "labridae", "pomacentridae", "tetraodontidae")
pretty_names <- c("Acanthuridae", "Apogonidae", "Balistoidae", "Chaetodontidae", "Labridae", "Pomacentridae", "Tetraodontidae")
names(fams) <- pretty_names

tre_trait <- lapply(paste0(base, "input_trees/", fams, ".tre"), read.tree)
tre_speciation <- lapply(paste0(base, fams, ".tre"), read.tree)
names(tre_trait) <- names(tre_speciation) <- fams

cl <- makePSOCKcluster(detectCores(), useXDR=F)
clusterExport(cl, list("base", "fams", "tre_speciation", "tre_trait"))
evt_speciation <- parLapplyLB(cl, fams, function(fam) {
  tre <- tre_speciation[[fam]]
  BAMMtools::getEventData(tre, paste0(base, "div_runs/", fam, "_event_data.txt"), burnin=0.1)
})
evt_trait <- parLapplyLB(cl, fams, function(fam) {
  tre <- tre_trait[[fam]]
  BAMMtools::getEventData(tre, paste0(base, "trait_runs/", fam, "_PC1_event_data.txt"), burnin=0.1, type="trait")
})
names(evt_trait) <- names(evt_speciation) <- fams
clusterExport(cl, list("evt_speciation", "evt_trait"))
rtt_speciation <- parLapplyLB(cl, fams, function(fam) {
  BAMMtools::getRateThroughTimeMatrix(evt_speciation[[fam]])
})
rtt_trait <- parLapplyLB(cl, fams, function(fam) {
  BAMMtools::getRateThroughTimeMatrix(evt_trait[[fam]])
})
prior_speciation <- parLapplyLB(cl, fams, function(fam) {
  BAMMtools::getBranchShiftPriors(tre_speciation[[fam]], paste0(base, "div_runs/", fam, "_prior_probs.txt"))
})
prior_trait <- parLapplyLB(cl, fams, function(fam) {
  BAMMtools::getBranchShiftPriors(tre_trait[[fam]], paste0(base, "trait_runs/", fam, "_PC1_prior_probs.txt"))
})

names(pretty_names) <- names(prior_trait) <- names(prior_speciation) <- names(rtt_trait) <- names(rtt_speciation) <- fams
stopCluster(cl)
```

```{r bamm_plots, message=FALSE, echo=FALSE, warning=FALSE, results="hide", fig.show="hold", anchor="figure", fig.cap="Rates of shape evolution for PC1 (a, b) and speciation (c, d) across seven families of fishes. Phylorate plots (a, c) color branch lengths by rates of shape evolution (a) and speciation (c), where warmer colors indicate faster rates of evolution. Significant rate shift events (pp > 0.95) are indicated on the phylorate plot as a red circle on the corresponding branch. Median log rates of shape evolution (b) and speciation (d) through time, where black lines indicate the background rate and red lines indicate the rate of evolution in a clade experiencing a significant shift in rate, corresponding to red circles in (a) or (c).", fig.height=9.5, out.height="0.9\\textheight", out.width="\\textwidth"}

layout_mat <- matrix(c(
  0,  1,  2,  3,  4,
  5,  12, 13, 14, 15,
  6,  16, 17, 18, 19,
  7,  20, 21, 22, 23,
  8,  24, 25, 26, 27,
  9,  28, 29, 30, 31,
  10, 32, 33, 34, 35,
  11, 36, 37, 38, 39
), ncol = 5, byrow = TRUE)

@@

plot_rtt <- function(rtt, shifts = list(), xlim=NULL, ylim=NULL) {
  vartype <- if (rtt$type == "trait") "beta" else "lambda"
  if (length(shifts) > 0) {
    allshifts <- c(list(rtt), shifts)
  } else {
    allshifts <- list(rtt)
  }
  allshifts <- lapply(allshifts, function (x) {
    x$avg <- apply(x[[vartype]], 2, median)
    x$times <- rev(x$times)
    x
  })
  xlim <- if (is.null(xlim)) range(sapply(allshifts, `[`, "times")) else xlim
  ylim <- if (is.null(ylim)) range(sapply(allshifts, `[`, "avg")) else ylim
  old.par <- par(mar = c(2,3,2,1))
  plot.new()
  plot.window(rev(xlim), ylim, log = "y")
  axis(at = axTicks(1), side = 1, labels = axTicks(1))
  axis(side = 2, las = 2)
  lines(allshifts[[1]]$times, allshifts[[1]]$avg)
  if (length(shifts) > 0) {
    lapply(2:length(allshifts), function (x) {
      lines(allshifts[[x]]$times - min(allshifts[[x]]$times), allshifts[[x]]$avg, col="red")
    })
  }
  par(old.par)
  return(invisible(NULL))
}

mar <- c(2,1,2,1)
par(mar = c(0,0,0,0))
layout(layout_mat, widths = c(1, 6, 6, 6, 6), heights = c(0.9, 9, 9, 9, 9, 9, 9, 9))
htext("Shape Phylorates")
text(0, 0.5, "a)", cex=1.2)
htext("Shape Rate Through Time")
text(0, 0.5, "b)", cex=1.2)
htext("Speciation Phylorates")
text(0, 0.5, "c)", cex=1.2)
htext("Speciation Rate Through Time")
text(0, 0.5, "d)", cex=1.2)
lapply(pretty_names, vtext)
par(mar = mar)
lapply(fams, function(fam) {
  interval <- if(is.null(customintervals[[fam]])) c(0, 0.001) else customintervals[[fam]] 
  q <- plot.bammdata(evt_trait[[fam]], logcolor = F, breaksmethod = "jenks", color.interval = interval)
  best <- getBestShiftConfiguration(evt_trait[[fam]], prior = prior_trait[[fam]])
  addBAMMshifts(best, cex = 2, par.reset=F)
  axisPhylo()

  shift_node <- getShiftNodesFromIndex(best, 1)
  if (length(shift_node) > 0) {
    background_rates <- getRateThroughTimeMatrix(evt_trait[[fam]], node = shift_node, nodetype = "exclude")
    clade_rates <- lapply(shift_node, function (x) {
      phy <- extract.clade(tre_trait[[fam]], x)
      write.tree(phy, file = paste0(fam, "_trait_shift_", x, ".tre"))
      getRateThroughTimeMatrix(evt_trait[[fam]], node = x)
      })
    plot_rtt(background_rates, clade_rates)
  } else {
    plot_rtt(rtt_trait[[fam]])
  }

  q <- plot.bammdata(evt_speciation[[fam]], logcolor = F, breaksmethod = "jenks", color.interval = )
  best <- getBestShiftConfiguration(evt_speciation[[fam]], prior = prior_speciation[[fam]])
  addBAMMshifts(best, cex = 2, par.reset=F)
  axisPhylo()

  shift_node <- getShiftNodesFromIndex(best, 1)
  if (length(shift_node) > 0) {
    background_rates <- getRateThroughTimeMatrix(evt_speciation[[fam]], node = shift_node, nodetype = "exclude")
    clade_rates <- lapply(shift_node, function (x) {
      phy <- extract.clade(tre_speciation[[fam]], x)
      write.tree(phy, file = paste0(fam, "_speciation_shift_", x, ".tre"))
      getRateThroughTimeMatrix(evt_speciation[[fam]], node = x)
      })
    plot_rtt(background_rates, clade_rates)
  } else {
    plot_rtt(rtt_speciation[[fam]])
  }
})

@@

Using the broken-stick method of determining a PCA stopping point, we analyzed PC 1 through PC `r stopping_pc`. We project per-species consensus shapes into Procrustes space (Figure `r figr("sevenfamily_tangent_plot")`, Supplemental Figure S5). The BAMMtools analysis uncovered substantial amounts of heterogeneity in the rate of body shape evolution and speciation in each family (Figure `r figr("bamm_plots")`). Significant shifts in the rate of shape evolution or speciation were detected in three families: Labridae, Apogonidae, and Pomacentridae. The significant shifts in speciation rate corroborate those found in @Cowman2011 through either MEDUSA [@Alfaro2009] or a relative cladogenesis statistic [@Nee1992]. Two significant shifts in shape evolution rate occur in the wrasses (Labridae). The first rate shift occurs deep in the tree, corresponding to the lineage containing the labrine, scarine, and cheiline tribes. The other shift is nested within that group, in *Sparisoma*.  One shift in speciation rate also occurs in the wrasses, encompassing the genera *Chlorurus* and *Scarus*. One shift in speciation rate occurs in the cardinalfishes (Apogonidae), encompassing members of the genera *Apogon*, *Archamia*, *Zoramia*, *Ostorhinchus*, *Cheilodpterus*, *Gossamia*, *Fowleria*, and *Phaeoptyx* [Apogonini + Apogonichthynini *sensu* @Mabuchi2014]. One shift in the rate of shape evolution occurs in the damelfishes (Pomacentridae) in the genus *Amphiprion*.

>>>>

```{r prelude, echo=FALSE, message=FALSE, warning=FALSE}
?library("MASS")
library("readr")
library("dplyr")
library("ggplot2")
library("stringr")
library("magrittr")
library("tidyr")
library("reshape2")
library("scales")
library("geomorph")
library("rjson")
library("ipred")
library("vegan")
library("lubridate")
library("directlabels")
library("BAMMtools")
library("knitr")
library("kfigr")
library("knitcitations")
library("parallel")
library("fish.reliability")
options("citation_format"="pandoc")
cleanbib()
set.seed(111) # reproducible and stable sampling
data(fish_reliability)
data(fish_families)
knitr::opts_chunk$set(echo=FALSE, cache=1, fig.align="center")
options(mc.cores = parallel::detectCores())
```

@@

Although it is now possible to build phylogenetic trees with thousands of tips, and phenotypic data sets have similarly been growing larger and larger, the traits that are typically studied at this scale tend to be simple: geographic occurrences [@Jetz2012], one or two continuous characters [@Harmon2010; @Rabosky2013], a single discrete character [@Goldberg2010; @Edwards2011; @Price2012], or some combination of these [@Zanne2014; @Pyron2014]. A richer understanding of the forces that shape macroevolution requires the collection of more detailed phenotypic trait data at a broad scale.

@@

Collecting landmark-based geometric morphometric data at a broad scale permits detailed analysis of different sources of error, such as among- and within-observer variation [@VonCramon-Taubadel2007]. To assess whether the quality of data gathered by workers recruited through Amazon Mechanical Turk was significantly different than traditionally-collected data, we asked turkers ($n = `r nrow(filter(fish_reliability, role == "turker") %>% distinct(sha1mac))`$) and experts ($n = `r nrow(filter(fish_reliability, role == "morphologist") %>% distinct(sha1mac))`$) to landmark a set of five fish images, five times each. All participants used the same protocol and same software to digitize the same set of fishes. The landmarks were carefully selected based on previously-published literature concerning fish shape [Supplemental Figure S2; @Fink1995; @Cavalcanti1999; @Ruber2001a; @Klingenberg2003; @Chakrabarty2005; @Frederich2008; @Claverie2014; @Thacker2014]. We also ensured that the chosen landmarks included morphological features that were relatively straightforward to digitize (e.g., the position of the eye) and features that were likely to be more challenging to digitize (e.g., the most anterior and most dorsal points of the preopercle), in order to test for turker and expert differences over a spectrum of difficulties. We report the inter-observer reliability for turkers and experts by computing the ratio of the among-individual and the sum of the among-individual and measurement error variance components in a repeated measures nested MANOVA [@Palmer1986; @Zelditch2012].

To assess the differences between turker and experts on a per-landmark basis, we first compared for each landmark the median position of all turkers to the median position of all experts. We assumed that the expert median was the true position of that landmark, and calculated the absolute Euclidian distance in pixels. Larger distances would indicate low turker accuracy, while smaller distances would indicate high turker accuracy. We then examined the variance in turker landmarks. For each landmark, we rotated the cloud of points to maximize variance in one dimension, and calculated the log-ratio of median absolute deviations (MAD) between turkers and experts. This rotation is a conservative approach for assessing the difference in variance between these two groups, because it maximizes any apparent differences in landmark position. A positive log-ratio indicated that experts had lower variance than turkers, while a negative log-ratio indicated that turkers had lower variance. For all subsequent analysis, we excluded landmarks where turkers performed especially poorly, where either the accuracy or precision components for a given landmark exceeded 1.5 times the interquartile range of that component.

@@

To demonstrate the utility of obtaining comparative data using this method, we use previously published phylogenies for seven fish families: Acanthuridae [@Sorenson2013], Balistoidae, Tetraodontidae [@Santini2013], Apogonidae, Chaetodontidae, Labridae [@Cowman2011; @Choat2012], and Pomacentridae [@Frederich2013]. We matched `r nrow(distinct(fish_families, tip))` species to left-lateral images from the Encyclopedia of Life (<http://eol.org/>) using their application programming interface [Supplemental Table S1; @Parr2014]. Crowdsourced workers placed landmarks describing body shape variation following a standard protocol (Supplementary Material). The Cartesian position of these landmarks were used in a generalized Procrustes analyses [@Gower1975; @Rohlf1990], which centers, scales, and rotates landmark configurations to minimize the least-squares distance between shapes. We then determined the major components of shape variation using a Procrustes-aligned principal components analysis (PCA) [@Mardia1979; @Bookstein1991] with the R package *geomorph* `r citep(citation("geomorph"))`, and retain the principal component axes whose eigenvalues exceeded the corresponding random broken-stick component [@Jackson1993; @Legendre1998] for all subsequent analyses.

We used Bayesian Analysis of Macroevolutionary Mixtures [BAMM; @Rabosky2014a] to estimate rates of body shape evolution for all seven families. BAMM estimates the location of rate shifts in character evolution using a transdimensional (reversible jump) Markov Chain Monte Carlo method that samples a variety of models of lineage diversification and trait evolution. We assessed convergence and mixing using Tracer [@rambaut2007tracer]. We also repeated each analysis and simulated under the prior (without data) to exclude rate heterogeneity that occurred solely due to stochastic processes. We use a Bayes Factor criterion of $BF > 5$ to enumerate the set of credible shifts [@Shi2015] and visualized them in R using BAMMtools `r citep(citation("BAMMtools"))`.

@@

The inter-observer reliability of turkers and experts as measured by the ratio of among-individual and sum of the among-individual and measurement error ANOVA components was `r percent(reliability$turker)` and `r percent(reliability$expert)`, respectively. Although there is no current standard for acceptable levels of measurement reliability [@VonCramon-Taubadel2007], these percentages are not low enough to suggest weaknesses in the measurement protocol.

@@

We projected turker and expert shape configurations into morphospace (Figure `r figr("turk_morph_procrustes_space")`, Supplemental Figure S4) Although the overall space occupied by each family's shape configurations vary, the aggregated median turker and expert shapes are not qualitatively different. The only exception is the triggerfishes (Balistidae), likely due to turker confusion over the exact location of dorsal fin due to their reduced anterior dorsal fin.

@@

Turkers in aggregate required `r as.character(total_turker_time)` to complete all `r nrow(distinct(fish_families, tip))` images at 5x replication `r figr("timing_plot")`. We estimate that a single expert would need `r as.character(morpho_per_image * nrow(distinct(fish_families, tip)))` to complete all images at 1x replication, extrapolated from a median expert time per image of `r as.character(morpho_per_image)`. At 5x replication, our projected expert would instead need `r as.character(morpho_per_image * nrow(fish_families))`.

@@

```{r}

par_get_rtt_matrix <- function(ephy, start.time=NULL, end.time=NULL, nslices=100, node=NULL, nodetype = 'include', mc.cores = getOption("mc.cores", 2L)) {

	if (!'bammdata' %in% class(ephy)) {
		stop("Object ephy must be of class 'bammdata'\n");
	}
  if (ephy$type == 'diversification') {
	  stop("'diversification' not supported")
	}

	if (is.null(node)) {
		nodeset <- ephy$edge[,2];
	} else if (!is.null(node) & nodetype == 'include') {
		#nodeset <- getDesc(ephy, node)$desc_set;
		nodeset <- unlist(sapply(node, function(x) getDesc(ephy, x)$desc_set))
	} else if (!is.null(node) & nodetype == 'exclude') {
		nodeset <- setdiff( ephy$edge[,2], unlist(sapply(node, function(x) getDesc(ephy, x)$desc_set)));
	} else {
		stop('error in getRateThroughTimeMatrix\n');
	}

	bt <- branching.times(as.phylo.bammdata(ephy));
	maxpossible <- max(bt[as.character(intersect(nodeset, ephy$edge[,1]))]);

	if (is.null(start.time)) {
		start.time <- max(bt) - maxpossible;
	}
	if (is.null(end.time)) {
		end.time <- max(bt);
	}

	decimals = function(x){
	if(x%%1 != 0)	{
		return(nchar(strsplit(as.character(x),".",fixed=TRUE)[[1]][[2]]));
	}		else	{
		return(10);
	}
}

	tvec <- seq(start.time, end.time, length.out= nslices);
	tol = 1*10^-decimals(ephy$eventBranchSegs[[1]][1,2]);
	mm <- matrix(NA, nrow=length(ephy$eventBranchSegs), ncol=length(tvec));

	# this is much faster than 'safeCompare'.
	goodTime <- function (vec, val, tol) {
		(vec[,2] - val <= tol) & (val - vec[,3] <= tol);
	}

	res <- mclapply(mc.cores = mc.cores, 1:length(ephy$eventBranchSegs), function(i) {
	  	es <- ephy$eventBranchSegs[[i]];
		events <- ephy$eventData[[i]];

		isGoodNode <- rep(TRUE, nrow(es));

		sapply(1:length(tvec), function(k) {
			isGoodTime <- goodTime(es, tvec[k], tol=tol);
			if (!(is.null(node))) { # only enter this if not the root. otherwise, only have to set once per i.
				isGoodNode <- es[,1] %in% nodeset;	
			}

			estemp <- es[isGoodTime & isGoodNode, ];
			tvv <- tvec[k] - events$time[estemp[,4]];
			rates <- exponentialRate(tvv, events$lam1[estemp[,4]], events$lam2[estemp[,4]]);
			mean(rates);
		})
	})

	mm <- do.call(rbind, res)

	obj <- list();
	obj$beta <- mm;
	obj$times <- tvec;

	class(obj) <- 'bamm-ratematrix';
	obj$type = 'trait';	
	return(obj);
}

```{r load_bamm_data, message=FALSE, echo=FALSE, warning=FALSE, results="hide"}
#base <- "data-raw/bamm/"
base <- "../data-raw/bamm/"
fams <- c("acanthurid", "apogonidae", "balistoidae", "chaetodontidae", "labridae", "pomacentridae", "tetraodontidae")
pretty_names <- c("Acanthuridae", "Apogonidae", "Balistoidae", "Chaetodontidae", "Labridae", "Pomacentridae", "Tetraodontidae")
names(fams) <- pretty_names

tre_trait <- lapply(paste0(base, "input_trees/", fams, ".tre"), read.tree)
names(tre_trait) <- fams

cl <- makePSOCKcluster(detectCores(), useXDR=F)
clusterExport(cl, list("base", "fams", "tre_trait"))
evt_trait <- parLapplyLB(cl, fams, function(fam) {
  tre <- tre_trait[[fam]]
  BAMMtools::getEventData(tre, paste0(base, "trait_runs/", fam, "_PC1_event_data.txt"), burnin=0.1, type="trait")
})
names(evt_trait) <- fams
clusterExport(cl, list("evt_trait"))
rtt_trait <- parLapplyLB(cl, fams, function(fam) {
  BAMMtools::getRateThroughTimeMatrix(evt_trait[[fam]])
})
prior_trait <- parLapplyLB(cl, fams, function(fam) {
  BAMMtools::getBranchShiftPriors(tre_trait[[fam]], paste0(base, "trait_runs/", fam, "_PC1_prior_probs.txt"))
})

names(pretty_names) <- names(prior_trait) <- names(rtt_trait) <- fams
stopCluster(cl)
```

```{r bamm_plots, message=FALSE, echo=FALSE, warning=FALSE, results="hide", fig.show="hold", anchor="figure", fig.cap="Rates of shape evolution for PC1 across seven families of fishes. (a) Phylorate plots color branch lengths by rates of shape evolution, where warmer colors indicate faster rates of evolution. Significant rate shift events (pp > 0.95) are indicated on the phylorate plot as a red circle on the corresponding branch. (b) Median log rates of shape evolution through time, where black lines indicate the background rate and red lines indicate the rate of phenotypic evolution in a clade experiencing a significant shift in rate, corresponding to red circles in (a).", fig.width = 6, fig.height = 3, out.height="3in", out.width="6in"}

@@

plot_rtt <- function(rtt, shifts = list(), xlim=NULL, ylim=NULL, labels = TRUE) {
  vartype <- if (rtt$type == "trait") "beta" else "lambda"
  if (length(shifts) > 0) {
    allshifts <- c(list(rtt), shifts)
  } else {
    allshifts <- list(rtt)
  }
  allshifts <- lapply(allshifts, function (x) {
    x$avg <- apply(x[[vartype]], 2, median)
    x$times <- rev(x$times)
    x
  })
  xlim <- if (is.null(xlim)) c(0, max(sapply(allshifts, `[[`, "times"))) else xlim
  ylim <- if (is.null(ylim)) range(sapply(allshifts, `[[`, "avg")) else ylim
  plot.new()
  plot.window(rev(xlim), ylim, log = "y")
  axis(at = axTicks(1), side = 1, labels = axTicks(1))
  axis(at = c(1e-4, 1e-3, 1e-2, 1e-1), side = 2, las = 2, labels = labels)
  lines(allshifts[[1]]$times, allshifts[[1]]$avg)
  if (length(shifts) > 0) {
    lapply(2:length(allshifts), function (x) {
      lines(allshifts[[x]]$times - min(allshifts[[x]]$times), allshifts[[x]]$avg, col="red", lwd=1.5)
    })
  }
  return(invisible(NULL))
}

mar <- c(1.5,1,0,0)+0.1
par(mar = c(0,0,0,0))

layout_mat <- matrix(c(
  8, 10, 12, 14, 16, 18, 20, 22, # shape phylorates
  9, 11, 13, 15, 17, 19, 21, 23, # shape rate through time
  0, 1, 2, 3, 4, 5, 6, 7
), nrow = 3, byrow = TRUE)

layout(layout_mat, widths = c(1.3, 12, 9, 9, 9, 9, 9, 9), heights = c(5, 6, 1))
lapply(pretty_names, htext) # family names
vtext("(a) Shape Phylorates")
vtext("(b) Shape Rate Through Time")
par(mar = mar)

ctr <- 1
lapply(fams, function(fam) {
  par(mar=mar)
  if (ctr == 1) {
    par(mar=c(1.5,4,0,0)+0.1)
  }
  interval <- if(is.null(customintervals[[fam]])) c(0, 0.001) else customintervals[[fam]] 
  q <- plot.bammdata(evt_trait[[fam]], logcolor = F, breaksmethod = "jenks", color.interval = interval)
  best <- getBestShiftConfiguration(evt_trait[[fam]], prior = prior_trait[[fam]])
  addBAMMshifts(best, cex = 2, par.reset=F)
  #axis_phylo(label_ticks = FALSE)
  axisPhylo(col.axis = "transparent")
  shift_node <- getShiftNodesFromIndex(best, 1)
  if (length(shift_node) > 0) {
    background_rates <- par_get_rtt_matrix(evt_trait[[fam]], node = shift_node, nodetype = "exclude", start.time=0)
    clade_rates <- lapply(shift_node, function (x) {
      phy <- extract.clade(tre_trait[[fam]], x)
      write.tree(phy, file = paste0(fam, "_trait_shift_", x, ".tre"))
      par_get_rtt_matrix(evt_trait[[fam]], node = x)
      })
    plot_rtt(background_rates, clade_rates, ylim=c(4e-5, 0.1), labels = ctr == 1)
  } else {
    plot_rtt(rtt_trait[[fam]], ylim=c(4e-5, 0.1), labels = ctr == 1)
  }
  ctr <<- ctr + 1
})

@@

Using the broken-stick method of determining a PCA stopping point, we analyzed PC 1 through PC `r stopping_pc`. We project per-species consensus shapes into Procrustes space (Figure `r figr("sevenfamily_tangent_plot")`, Supplemental Figure S5). The BAMMtools analysis uncovered heterogeneity in the rate of body shape evolution in each family (Figure `r figr("bamm_plots")`). Significant shifts in the rate of shape evolution were detected within two families: Labridae and Pomacentridae. Two significant shifts in shape evolution rate occur in the wrasses (Labridae). The first rate shift occurs deep in the tree, corresponding to the lineage containing the labrine, scarine, and cheiline tribes. The other shift is nested within that group, in *Sparisoma*. One shift in the rate of shape evolution occurs in the damelfishes (Pomacentridae) in the genus *Amphiprion*.

### 2015-09-22 manuscript/supplement.Rmd
Code drop

---
title: Supplementary material for "Crowdsourced geometric morphometrics enable rapid large-scale collection and
  analysis of phenotypic data"
author:
- affiliation: ucla
  email: jonathan.chang\@ucla.edu
  footnote: Author for correspondence jonathan.chang\@ucla.edu
  name: Jonathan Chang
- affiliation: ucla
  name: Michael E. Alfaro
bibliography:
- library.bib
- knitcitations.bib
output:
  pdf_document:
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    keep_tex: yes
    latex_engine: xelatex
    template: inc/mstemplate.tex
  html_document:
    toc: yes
  word_document: default
documentclass: inc/msarticle
endfloat: yes
fontsize: 11pt
geometry: margin=1.2in
layout: preprint
lineno: no
linestretch: 2
csl: inc/methods-in-ecology-and-evolution.csl
address:
- address: |
    Department of Ecology and Evolutionary Biology, University of California, Los Angeles, CA, USA
  code: ucla
---

@@

\renewcommand\thefigure{S\arabic{figure}}
\setcounter{figure}{0}

@@

@@

>>>>

---
title: Supplementary material for "Crowdsourced geometric morphometrics enable rapid large-scale collection and
  analysis of phenotypic data"
author:
- affiliation: ucla
  email: jonathan.chang\@ucla.edu
  footnote: Author for correspondence jonathan.chang\@ucla.edu
  name: Jonathan Chang
- affiliation: ucla
  name: Michael E. Alfaro
bibliography:
- library.bib
- knitcitations.bib
output:
  pdf_document:
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    keep_tex: yes
    latex_engine: xelatex
    template: inc/mstemplate.tex
  html_document:
    toc: yes
  word_document: default
documentclass: inc/msarticle
endfloat: yes
fontsize: 10pt
geometry: margin=1in
layout: preprint
lineno: no
linestretch: 1
csl: inc/methods-in-ecology-and-evolution.csl
address:
- address: |
    Department of Ecology and Evolutionary Biology, University of California, Los Angeles, CA, USA
  code: ucla
---

@@

\renewcommand\thefigure{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}

@@

## Testing turker vs expert consistency

```{r geomorph_consistency, warning=FALSE}
res <- gpa_with_vectors(only_good_points, family, role, sha1mac, sequence)

family <- res$vectors$family
role <- res$vectors$role
repno <- as.integer(res$vectors$sequence) %% 5
sha1mac <- as.factor(res$vectors$sha1mac)

devtools::install_version("geomorph", version = "2.1.6", local=F, quick=T, dependencies=F, upgrade_dependencies=F, keep_source=T)

devtools::install("~/geomorph/", local=F, quick=T, dependencies=F, upgrade_dependencies=F, keep_source=T)

lm_result <- advanced.procD.lm(res$coords ~ family*role*sha1mac, ~family*sha1mac, iter=9)
?lm_p <- lm_result$P[2]
lm_Z <- lm_result$Z[2]
lm_F <- lm_result$F[2]
rm(family, role, repno, sha1mac)
```

@@

```{r fish_families_table, message=FALSE, warning=FALSE, as.is=T}
fish_families_urls <- read_csv("fish_families_urls.csv")
knitr::kable(select(fish_families_urls, -eol_guid), caption='Families, species names, and links to the images hosted on Encyclopedia of Life for the section "Example: a phenomic pipeline for comparative phylogenetic analysis".')
```
